Enhancing Forecast Reconciliation: A Study of Alternative Covariance Estimators

### 2025-05-07 research_proposal/research_proposal_draft.qmd
Refine Alternative Covariance Estimators section

We reconstruct estimator of $\boldsymbol{W}_h$ as $\hat{\boldsymbol{W}}_h = k_h \, g(\hat{\boldsymbol{W}}_1)$, and $g(.)$ is an estimator function of the unbiased sample covariance of in-sample one-step-ahead base forecast errors $\hat{\boldsymbol{W}}_1 = \frac{1}{T} \sum_{t=1}^{T} \hat{\boldsymbol{e}}_{t|t-1} \hat{\boldsymbol{e}}_{t|t-1}'$

@@

The proposed MinT approach by @Wickramasuriya2019-xq uses the shrinkage estimator from @Schafer2005-yw. \textcolor{red}{It guarantees positive definiteness and variance reduction for the covariance matrix, especially when $p>T$}. The shrinkage estimator is given by:

$$
\hat{\boldsymbol{W}}^{shr}_{1, D} = \lambda_D \hat{\boldsymbol{W}}_{1, D} + (1 - \lambda_D) \hat{\boldsymbol{W}}_1
$$

$\hat{\boldsymbol{W}}_{1, D}$ is a diagonal matrix comprising the diagonal entries of $\hat{\boldsymbol{W}}_1$. This approach will shrink the covariance matrix $\hat{\boldsymbol{W}}_1$ towards its diagonal version, meaning the off-diagonal elements are shrunk towards zero while the diagonal ones remain unchanged.

@Schafer2005-yw also proposed an optimal shrinkage intensity parameter $\lambda_D$ for this setting, assuming the variances are constant:

@@

where $\hat{r}_{ij}$ is the $ij$th element of $\hat{\boldsymbol{R}}_1$, the 1-step-ahead sample correlation matrix (obtained from $\hat{\boldsymbol{W}}_1$) to shrink it toward an identity matrix

The optimal lambda is obtained by minimising the $MSE(\hat{\boldsymbol{W}}_1) = Bias(\hat{\boldsymbol{W}}_1)^2 + Var(\hat{\boldsymbol{W}}_1)$. More specifically, we trade the unbiasedness of the sample covariance matrix for a lower variance. The objective function itself does not take into account any possible principal components structure in the data, and is not flexible enough since it shrinks all off-diagonal elements equally towards zeros.

@@

The NOVELIST (NOVEL Integration of the Sample and Thresholded Covariance) estimator, proposed by @Huang2019-ua, is currently the main focus of this research project. \textcolor{red}{It introduces adaptive sparsity, retaining strong correlations while discarding weak, noisy links. NOVELIST offers more flexibility than the shrinkage estimator, however, it does not guarantee positive definiteness.}

@@

$$
\hat{\boldsymbol{W}}^{N}_{1} = \lambda_{\delta} \hat{\boldsymbol{W}}_{1, \delta} + (1 - \lambda_{\delta}) \hat{\boldsymbol{W}}_1
$$

By convenient setting, we rewrite it as in correlation matrix:

@@

In this setting, $\hat{\boldsymbol{R}}_{1,\delta}$ is the thresholded sample correlation matrix, where each element is regularised by:

$$
\hat{r}_{1,ij}^\delta = \operatorname{sign}(\hat{r}_{1,ij}) \, (|\hat{r}_{1,ij}| - \delta)_+.
$$

For a given threshold $\delta$, @Huang2019-ua derived the optimal shrinkage intensity parameter $\lambda(\delta)$ using Ledoit-Wolf's lemma [@Ledoit2003-qv]. It can be computed as:

$$
\hat{\lambda}(\delta) = \frac{
  \sum_{i \neq j} \widehat{Var}(\hat{r}_{1,ij}) \; I(|\hat{r}_{1,ij}| \leq \delta)
} {
  \sum_{i \neq j} (\hat{r}_{1,ij} - \hat{r}_{1,ij}^\delta)^2
}
$$

On the other hand, the optimal threshold $\delta^*$ does not have a closed-form solution, and is typically obtained by executing a rolling-window cross-validation procedure. The idea is to, for each rolling-window set, loop through a range of threshold values, compute the corresponding $\hat{\lambda}(\delta)$ and $\hat{\boldsymbol{R}}^{N}_{1}$. Then we select the one that minimises the average out-of-sample reconciled forecast errors. Although it is not required to fit forecasting models multiple time, the cross-validation procedure is still computationally expensive as it computes the NOVELIST estimator and perform reconciliation for each threshold value. The formal algorithm is given in the @sec-novelist_cv Appendix.

\textcolor{red}{Note that when $\delta \in \bigl[ \text{max}_{ij}|\hat{r}_{1,ij}|, \; 1 \bigr]$, the NOVELIST estimator collapses to the shrinkage estimator, and when $\delta = 0$, it becomes the sample covariance matrix.}

@@

The POET (Principal Orthogonal complEment Thresholding) estimator, proposed by @Fan2013-jz, is another "sparse" + "non-sparse" covariance estimator. \textcolor{red}{It takes the latent factors into account, and is appealing when there are common drivers in the time series within the hierarchy.}

@@

where $\hat{\gamma}_k$ and $\hat{\boldsymbol{\xi}}_k$ are the $k$th eigenvalue and eigenvector of the sample covariance matrix. $T(.)$ is the thresholding function, which can be either soft-thresholding, hard-thresholding, or others.

@@

\textcolor{red}{This approach is best of both worlds, leveraging the strengths of both NOVELIST and POET. The PC-adjusted NOVELIST overcomes the shortcomings of the current shrinkage estimator, taking prominent PCs into account while also offers extra flexibility.} The idea is to apply the NOVELIST estimator to the orthogonal complement part $\hat{\boldsymbol{R}}_{1, K}$, and then add the principal components part back. The PC-adjusted NOVELIST estimator is formulated as:

@@

and $\hat{\boldsymbol{R}}^{N}_{1, K}$ is the NOVELIST estimator applied to the orthogonal complement part $\hat{\boldsymbol{R}}_{1, K}$. Similar to the NOVELIST estimator, $\hat{\boldsymbol{R}}^{N, K}_{1}$ is not guaranteed to be positive definite.

>>>>

We reconstruct the estimator of $\boldsymbol{W}_h$ as $\hat{\boldsymbol{W}}_h = k_h \, g(\hat{\boldsymbol{W}}_1)$, where $k_h > 0$. The function $g(.)$ is an estimator of the unbiased sample covariance matrix of the in-sample one-step-ahead base forecast errors $\hat{\boldsymbol{W}}_1 = \frac{1}{T} \sum_{t=1}^{T} \hat{\boldsymbol{e}}_{t|t-1} \hat{\boldsymbol{e}}_{t|t-1}'$, and $T$ is the length of series' time dimension.

@@

The proposed MinT approach by @Wickramasuriya2019-xq uses the shrinkage estimator from @Schafer2005-yw. It guarantees positive definiteness and variance reduction for the covariance matrix, especially when the total number of series $n>T$. The shrinkage estimator is given by:

$$
\hat{\boldsymbol{W}}^{shr}_{1} = \lambda_D \hat{\boldsymbol{W}}_{1, D} + (1 - \lambda_D) \hat{\boldsymbol{W}}_1,
$$

where $\hat{\boldsymbol{W}}_{1, D}$ is a diagonal matrix comprising the diagonal entries of $\hat{\boldsymbol{W}}_1$. We refer to any $\lambda \in [0,1]$ as the shrinkage intensity parameter, the subscript specifies which estimator it belongs to. This approach shrinks the covariance matrix $\hat{\boldsymbol{W}}_1$ towards its diagonal matrix, meaning the off-diagonal elements are shrunk towards zero while the diagonal ones remain unchanged.

@Schafer2005-yw also proposed an estimate of the optimal shrinkage intensity parameter $\lambda_D$:

@@

where $\hat{r}_{ij}$ is the $ij$th element of $\hat{\boldsymbol{R}}_1$, the 1-step-ahead sample correlation matrix (obtained from $\hat{\boldsymbol{W}}_1$).

The optimal $\lambda_D$ is obtained by minimising the mean squared error of $\hat{\boldsymbol{W}}_1$: $MSE(\hat{\boldsymbol{W}}_1) = Bias(\hat{\boldsymbol{W}}_1)^2 + Var(\hat{\boldsymbol{W}}_1)$. More specifically, we trade the unbiasedness of the sample covariance matrix for a lower variance. The objective function itself does not take into account any possible principal components structure in the data, and is not flexible enough since it shrinks all off-diagonal elements equally towards zeros.

- Screeplot of eigenvalues

@@

The NOVELIST (NOVEL Integration of the Sample and Thresholded Covariance) estimator, proposed by @Huang2019-ua, is currently the main focus of this research project. \textcolor{red}{It introduces adaptive sparsity, retaining strong correlations while discarding weak, noisy correlations. NOVELIST offers more flexibility than the shrinkage estimator, however, it does not guarantee to be positive definite.}

@@

$$
\hat{\boldsymbol{W}}^{N}_{1} = \lambda_{\delta} \hat{\boldsymbol{W}}_{1, \delta} + (1 - \lambda_{\delta}) \hat{\boldsymbol{W}}_1,
$$

where $\hat{\boldsymbol{W}}_{1, \delta}$ is the thresholded version of $\hat{\boldsymbol{W}}_{1}$. By convenient setting, we can rewrite it in terms of correlation:

@@

In this setting, $\hat{\boldsymbol{R}}_{1,\delta}$ is the thresholded correlation matrix, where each element is regularised by:

$$
\hat{r}_{1,ij}^\delta = \text{sign}(\hat{r}_{1,ij}) \, 
\text{max}(|\hat{r}_{1,ij}| - \delta, \; 0),
$$

where $\delta \in [0,1]$ is the threshold parameter. For a given threshold $\delta$, @Huang2019-ua derived an analytical expression for the optimal shrinkage intensity parameter $\lambda(\delta)$ using Ledoit-Wolf's lemma [@Ledoit2003-qv]. It can be computed as:

- \textcolor{red}{@Huang2019-ua mentioned Ledoit-Wolf's lemma}

$$
\hat{\lambda}(\delta) = \frac{
  \sum_{i \neq j} \widehat{Var}(\hat{r}_{1,ij}) \; \boldsymbol{1}(|\hat{r}_{1,ij}| \leq \delta)
} {
  \sum_{i \neq j} (\hat{r}_{1,ij} - \hat{r}_{1,ij}^\delta)^2
},
$$

where $\boldsymbol{1}(.)$ is the indicator function.

On the other hand, the optimal threshold $\delta^*$ does not have a closed-form solution, and is typically obtained by executing a rolling-window cross-validation procedure. The formal algorithm is given in the @sec-novelist_cv Appendix. Although it is not required to fit forecasting models multiple times, the cross-validation procedure is still computationally expensive as it computes the NOVELIST estimator and perform reconciliation for each threshold value.

<!-- The idea is to, for each rolling-window set, loop through a range of threshold values, compute the corresponding $\hat{\lambda}(\delta)$ and $\hat{\boldsymbol{R}}^{N}_{1}$. Then we select the one that minimises the average out-of-sample reconciled forecast errors -->

Note that when $\delta \in \bigl[ \text{max}_{i \neq j}|\hat{r}_{1,ij}|, \; 1 \bigr]$, the NOVELIST estimator collapses to the shrinkage estimator, and when $\delta = 0$, it becomes the sample covariance matrix.

@@

The POET (Principal Orthogonal complEment Thresholding) estimator, proposed by @Fan2013-jz, is another "sparse" + "non-sparse" covariance estimator. It takes the latent factors into account, and is appealing when there are common drivers in the time series within the hierarchy.

@@

where $\hat{\gamma}_k$ and $\hat{\boldsymbol{\xi}}_k$ are the $k$th largest eigenvalue and the corresponding eigenvector of the sample covariance matrix, respectively, and $T(.)$ is the thresholding function, which can be either soft-thresholding, hard-thresholding, or others.

@@

This approach is best of both worlds, leveraging the strengths of both NOVELIST and POET. The PC-adjusted (Principal-Component-adjusted) NOVELIST overcomes the shortcomings of the current shrinkage estimator, taking prominent PCs into account while also offers extra flexibility. The idea is to apply the NOVELIST estimator to the orthogonal complement part $\hat{\boldsymbol{R}}_{1, K}$, and then add the principal components part back. The PC-adjusted NOVELIST estimator is formulated as:

@@

where $\hat{\boldsymbol{R}}^{N}_{1, K}$ is the NOVELIST estimator applied to the orthogonal complement part $\hat{\boldsymbol{R}}_{1, K}$. Similar to the NOVELIST estimator, $\hat{\boldsymbol{R}}^{N, K}_{1}$ is not guaranteed to be positive definite.

Methods to ensure positive definiteness of the NOVELIST estimator (and its PC-adjusted variant) will be explored and studied in the project. @Huang2019-ua proposed to diagonalise the NOVELIST
estimator and replace any eigenvalues that fall under a certain small positive threshold by the value of that threshold. Alternatively, we can implement the algorithm of @Higham2002-te that computes the nearest positive definite matrix to a given matrix.
